\section{Related Work}
\paragraph{Type-based numerical analysis techniques.}
This work extends Numerical Fuzz \cite{numfuzz}, which is based off the Fuzz line
of work \cite{fuzz} that uses sensitivities and monads for reasoning about
differential privacy. Relatedly, Bean \cite{bean} is a typed-based approach to
backwards error analysis, which is a different notion of error than in Numerical
and Negative Fuzz.
A floating-point program is backwards stable if the floating-point result is the
correct result of running the ideal program given approximate nearby inputs.
Backwards error analysis seeks to bound the maximum distance between the actual
and approximate inputs.
Martel \cite{martel} uses a dependent type system to track forwards error.
However, Martel proves a weaker subject reduction theorem and does not prove an
error soundness theorem.
% TODO: compare martel here

\paragraph{Other automated numerical analysis techniques.}
Fluctuat \cite{fluctuat} and Gappa \cite{gappa} both use interval
analysis and can reason about programs with conditionals and loops. 
FPTaylor \cite{fptaylor} uses symbolic taylor expansions and global optimization
to estimate round-off error.
Abbasi and Darulova \cite{abbasi-darulova} also use taylor expansions but take
a more modular approach that allows analyses to be shared at different function
call sites.
Rosa \cite{rosa1} \cite{rosa2} also uses taylor expansions for handling
propagation error.
By contrast, our approach supports different rounding modes and higher-order
functions. However, we do not support recursion and restrict conditionals from
casing on floating-point computations.

\paragraph{Interval-style bounds analysis in the type system.}
The Checker framework \cite{checker} provides a modular approach to extending
Java's type system with new typing rules. Several interval analysis checkers
have been written \cite{checker-whole-interval} \cite{checker-modular-interval},
but they are designed to find bugs rather than for a roundoff error
analysis.
There is also work to study type inference from the perspective of a generic
abstract interpretation analysis \cite{type-to-abstract-cousot} and to be able
to turn an abstract interpretation analysis into a type inference problem
\cite{abstract-to-type}. 

% \paragraph{Grothendieck construction of the integers from the naturals.}
% The idea of constructing the integers via a pairing of the naturals has existed
% since at least Dedekind's work in 1872.
% The construction also appears in several introductory textbooks in algebra and
% category theory.

% A major challenge in the automated numerical analysis literature is scalability.
% Many existing approaches rely on global optimization \cite{fptaylor}
% \cite{satire}, rewrite saturation \cite{gappa}, or SMT-based methods
% \cite{rosa}. However, as programs scale, these analysis approaches frequently
% time-out. Recent work has applied typed-based analysis approaches to error
% analysis, such as Numerical Fuzz \cite{numfuzz} and Bean \cite{bean}. There is
% no need to perform global optimization, run an algorithm to convergence,
% saturate rewrites, or bit-blast.
%
