\section{Encoding} \label{sec:encoding}
In this section we encode Numerical Fuzz to provide \textit{a priori} and
\textit{a posterori} error bounds in the presence of subtraction and negative
numbers.
In Section~\ref{sec:instantiation}, we instantiate Numerical Fuzz using the
modular interface defined in Definition~\ref{def:numfuzz-interface} using a
$\textit{paired representation}$ introduced below. Then, in
Section~\ref{sec:application} we demonstrate how to read out the
$\textit{paired-representation}$ as concrete bounds for standard, unpaired,
floating-point arithmetic.

\subsection{Instantiation} \label{sec:instantiation}
Our forwards error analysis relies on bounding Lipschitz function sensitivity.
However, subtraction, and by extension addition with negative numbers, has
unbounded Lipschitz sensitivity over the reals. This poses a challenge to a
compositional type-based analysis of programs involving subtraction and negative
numbers.

% TODO: do not repeat, the below section got moved to technical overview
% To avoid this problem, we encode our semantics through a \textit{paired
% representation} where we associate for each real $r \in \mathbb{R}$ a pair $(r,
% a, b) \in \mathbb{P} = \mathbb{R} \times \mathbb{R}^+ \times \mathbb{R}^+ = \mathit{num}$ such that $r = a -
% b$. 
We measure distance in $\mathbb{P}$ via the following function: $d_\mathbb{P}((r, a, b), (r, a', b'))
\triangleq max(d_\mathbb{R}(a,a'), d_\mathbb{R}(b,b'))$.

\begin{lemma}[$\textit{num}$ forms a metric space.]
\end{lemma}
\begin{proof}
Symmetry and reflexivity hold trivially. Triangle inequality holds from triangle
inequality on $d_{\mathbb{R}}$.
\end{proof}

% To avoid this problem, we encode our semantics through a \textit{paired
% representation} where we associate for each real $r \in \mathbb{R}$ a pair $(a,
% b) \in \mathbb{R}^2$ such that $r = a - b$. 
Our operations and interval bounds operate over the paired representation $\mathbb{P}$.
In this manner, subtraction and negative numbers can be encoded using bounded-sensitivity operations over the non-negative reals, avoiding the problem of infinite sensitivity.
Each of our operations $\Sigma = \{ \textbf{add}, \ \textbf{add}, \ \textbf{mul}
\}$ and $\Sigma_\textbf{num} = \{ +, -, \cdot \}$ are defined as follows:

\begin{equation}
\begin{aligned}[c]
  &\textbf{add} : \forall \epsilon_1, \epsilon_2, \mathbf{num}_{\epsilon_1} \times
  \mathbf{num}_{\epsilon_2} \multimap \mathbf{num}_{\epsilon_1 + \epsilon_2} \\
  &\textbf{add} \ \bnd{\{\epsilon_1\}} \ \bnd{\{\epsilon_2\}} \ (\langle (r, a,
  b), (r', a', b') \rangle) \mapsto (r + r', a + a', b + b') \\
  \\
  &\textbf{sub} : \forall \epsilon_1, \epsilon_2, \mathbf{num}_{\epsilon_1} \times
  \mathbf{num}_{\epsilon_2} \multimap \mathbf{num}_{\epsilon_1 - \epsilon_2} \\
  &\textbf{sub} \ \bnd{\{\epsilon_1\}} \ \bnd{\{\epsilon_2\}} \ (\langle (r, a, b),
  (r, a', b') \rangle ) \mapsto (r - r', a + b', b + a') \\
  \\
  &\textbf{mul} : \forall \epsilon_1, \epsilon_2, \mathbf{num}_{\epsilon_1} \times
  \mathbf{num}_{\epsilon_2} \multimap \mathbf{num}_{\epsilon_1 \cdot \epsilon_2} \\
  &\textbf{mul} \ \bnd{\{\epsilon_1\}} \ \bnd{\{\epsilon_2\}} \ (((r, a, b), (r, a',
  b'))) \mapsto (r * r', a * b + b * b', b * a' + a * b') \\
\end{aligned}
\end{equation}

Given a rounding function $\rho_\mathbb{R}$ over the reals, we can also
associate a paired $\rho_{\mathbb{P}}((r, a, b)) = (\rho_{\mathbb{R}}(r), a
\cdot \frac{\rho_{\mathbb{R}}(r)}{r}, b \cdot \frac{\rho_{\mathbb{R}}(r)}{r})$. 
It is sound to use the same round-off error bound $u$ associated with $\rho_\mathbb{R}$ for $\rho_\mathbb{P}$:

\begin{lemma}[Property of rounding and constant parameter $u$]
We wish to show that given a $u$ and $\rho_{\mathbb{R}}$ such that 
    $\forall e \in d_\mathbb{R} (\rho(e), e)
    \leq u$, then
    $\forall e \in \mathcal{R}_{num}, \mathcal{SD}_{\mathbf{num}}(\rho_{\mathbb{P}}(e), e)
    \leq u$. In other words, our translated rounding function from the unpaired representation respects the error bounds on the unpaired representation.
\end{lemma}
\begin{proof}
$(\rho_{\mathbb{P}}((r, a, b)), (r, a, b)) = 
(\rho_{\mathbb{R}}(r), a
\cdot \frac{\rho_{\mathbb{R}}(r)}{r}, b \cdot \frac{\rho_{\mathbb{R}}(r)}{r})$ So, 
$
d_{\mathbb{P}}((\rho_{\mathbb{R}}(r), a
\cdot \frac{\rho_{\mathbb{R}}(r)}{r}, b \cdot \frac{\rho_{\mathbb{R}}(r)}{r}), (r, a, b)) = max(\rho_{\mathbb{R}}(r), \rho_{\mathbb{R}}(r)) = \rho_{\mathbb{R}} (r)
$.
\end{proof}

Moreover, our operations satisfy metric preservation and our interval operations
are closed over intervals ($\textit{num}^2$).
% todo: these are closed over something other than num?

\begin{lemma}[Operations satisfy metric preservation]
Each $\textbf{op} : \tau \in \Sigma$ satisfies metric preservation.
\end{lemma}
\begin{proof}
Holds by unfolding.
\end{proof}


\begin{lemma}[Interval operations closure.]
Each interval operation is closed over our intervals $\textit{num}^2$.
\end{lemma}
\begin{proof}
Holds by inspection.
\end{proof}

By stapling together the proceeding lemmas, the interface ($\mathbb{P}, d_{\mathbb{P}}, \rho_{\mathbb{P}}, u, \Sigma, \Sigma_{\mathbf{num}}$) satisfy the interface for instantiating Numerical Fuzz specified in Definition~\ref{def:numfuzz-interface}.

\subsection{Application} \label{sec:application}
For a given $. \vdash e : M_q \textbf{num}_{(k_0, k_1)}$, we wish to bound the
error between $e$ computed ideally over the reals versus the floating-point approximation. We now prove our paired error theorems (Theorem~\ref{thm:paired-a-posteriori}, Theorem~\ref{thm:paired-a-priori}), which for a triple $(r, a, b)$ relates error on the paired components $a, b$ with error on $r$.

\begin{theorem}[Paired error theorem; a posteriori] \label{thm:paired-a-posteriori}
For $. \vdash e : M_q~\mathbf{num}_{(k_0, k_1)}$, $e \mapsto ((r, a, b), (r', a', b'))$ where $d_{\mathbb{R}}(r, r')$ is less than or equal to all of the following bounds: 
\begin{enumerate}
  \item $max(|ln(e^u + \frac{a}{r'}(1-e^{2u}))|,|ln(e^{-u} + \frac{a}{r'}(1-e^{-2u}))|)$
  \item $max(|ln(e^{-u} + \frac{b}{r'}(e^{-2u}- 1))|,|ln(e^u + \frac{b}{r'}(e^{2u}-1))|)$
\end{enumerate}
\end{theorem} 
\begin{proof}
  By our logical relation and metric preservation theorem (Theorem
  \ref{thm:metric-preservation}), we know that $e \mapsto ((r, a, b), (r', a', b'))$ such that $d_{\mathbb{P}}((r, a, b), (r', a', b')) \leq q$.
  Therefore, $d_{\mathbb{R}}(a, a'), d_{\mathbb{R}}(b, b') \leq q$.
  Unfolding the definition of $d_\mathbb{R}$, which uses relative error, we wish
  to bound $|ln(\frac{r}{r'})| = |ln(\frac{a - b}{a' - b'})|$.
  We proceed to prove the first bound:

  There are two cases, $r \leq r'$ or $r > r'$. In the first case, we wish to
  minimize $a-b$ and maximize $a' - b'$. So, 
  \begin{equation} \label{eq:a-priori-bnd-a-pos}
  \begin{aligned}[c]
    |ln(\frac{r}{r'})| \leq 
    |ln(\frac{a - b'e^u}{a' - b'})| =
    |ln(\frac{a - b' e^u + a'e^u - a'e^u}{a' - b'})| \\ =
    |ln(e^u + \frac{a - a'e^u}{a' - b'})| \leq
    |ln(e^u + \frac{a - ae^{2u}}{a' - b'})| =
    |ln(e^u + \frac{a}{r'}(1-e^{2u}))|
  \end{aligned}
  \end{equation}
  Similarly, for the other case, we have:
  \begin{equation} \label{eq:a-priori-bnd-a-neg}
  \begin{aligned}[c]
    |ln(\frac{r}{r'})| \leq 
    |ln(\frac{a - b'e^{-u}}{a' - b'})| =
    |ln(\frac{a - b'e^{-u} + a'e^{-u} - a'e^{-u}}{a' - b'})| \\ =
    |ln(e^{-u} + \frac{a - a'e^{-u}}{a' - b'})| \leq
    |ln(e^{-u} + \frac{a - ae^{-2u}}{a' - b'})| =
    |ln(e^{-u} + \frac{a}{r'}(1-e^{-2u}))|
  \end{aligned}[c]
  \end{equation}
  A sound bound for both cases is the maximum of Equation~\ref{eq:a-priori-bnd-a-pos} and Equation~\ref{eq:a-priori-bnd-a-neg}, which is exactly our first bound (1).
  The proof strategy for the second bound (2) mirrors that of the first bound.
\end{proof}

\begin{theorem}[Paired error theorem; a priori] \label{thm:paired-a-priori}
  For $. \vdash e : M_q~\mathbf{num}_{(k_0, k_1)}$, $e \mapsto ((r, a, b), (r', a', b'))$ where $d_{\mathbb{R}}(r, r')$ is less than or equal to all of the following bounds: 
\begin{enumerate}
  \item $max(|ln(e^{u} + \frac{a}{a-b}(e^{-u}- e^{u}))|, |ln(e^{-u} + \frac{a}{a-b}(e^{u}- e^{-u}))|)$
  \item $max(|ln(e^u + \frac{b}{a-b}(e^u - b e^{-u}))|,|ln(e^u + \frac{b}{a-b}(e^u - e^{-u}))|)$
\end{enumerate} 
\end{theorem} 
\begin{proof}
  Our error metric is symmetric so for this theorem we wish to prove a bound on
  $d_{\mathbb{R}}(r', r)$. Similarly to Theorem~\ref{thm:paired-a-posteriori},
  we prove each bound and, for each bound, split into two cases: $r \leq r'$ or
  $r > r'$. In the first case, we have:
  \begin{equation} \label{eq:a-priori-bnd-a-pos}
  \begin{aligned}[c]
    |ln(\frac{r'}{r})| \leq 
    |ln(\frac{a' - be^u}{a - b})| =
    |ln(\frac{a' - be^u + ae^u - ae^u}{a - b})| \\ =
    |ln(e^u + \frac{a' - ae^u}{a - b})| \leq
    |ln(e^u + \frac{ae^{-u} - ae^u}{a - b})| =
    |ln(e^u + \frac{a}{a - b}(e^{-u} - e^u))|
  \end{aligned}
  \end{equation}
  In the second case, we have:
  \begin{equation} \label{eq:a-priori-bnd-a-neg}
  \begin{aligned}[c]
    |ln(\frac{r'}{r})| \leq 
    |ln(\frac{a' - be^{-u}}{a - b})| =
    |ln(\frac{a' - be^{-u} + ae^{-u} - ae^{-u}}{a - b})| \\ =
    |ln(e^{-u} + \frac{a' - ae^{-u}}{a - b})| \leq
    |ln(e^{-u} + \frac{ae^{u} - ae^{-u}}{a - b})| =
    |ln(e^{-u} + \frac{a}{a - b}(e^{u} - e^{-u}))|
  \end{aligned}
  \end{equation}
  A sound bound for both cases is the maximum of Equation~\ref{eq:a-priori-bnd-a-pos} and Equation~\ref{eq:a-priori-bnd-a-neg}, which is exactly our first bound (1).
  The proof strategy for the second bound (2) mirrors that of the first bound.
\end{proof}

% TODO: detail how to plug in?

